---
title: "MSK-IMPACT mutation analysis"
description: |
  Clincally actionable mutations in MSK-IMPACT
author:
  - name: Larson Hogstrom 
    url: 
    affiliation: UiO
    affiliation_url: 
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
```

This document describes identifying clinically actionable mutaitons in the MSK-IMPACT variant set.

First, run the script to load the MSK data and compile the list of clinically actionable mutations

```{r, include=FALSE, echo=FALSE}
source("./CIViC_moa_curation.r")
```


### join with MOA - by protein change

```{r, include=TRUE, echo=TRUE}
dbAlteration$MOAset <- "Y"
msk.protein <- msk %>%
  dplyr::left_join(dbAlteration,by=c("Hugo_Symbol"="gene",
                            "protein_change"="AAChange")) 

msk.protein$actionability.summary.mod <- msk.protein$actionability.summary
msk.protein[grepl(";",msk.protein$actionability.summary),"actionability.summary.mod"] <- "multiple actionability types"
msk.protein[grepl("Predictive therapy actionability",msk.protein$actionability.summary) & grepl(";",msk.protein$actionability.summary),"actionability.summary.mod"] <- "Predictive therapy actionability or Prognistic"
```

```{r, include=TRUE, echo=TRUE}
#print(table(msk.protein$MOAset,msk.protein$source,exclude = NULL))

```



### rank mutations
Now rank order clinically actionable mutations by how often they occur in MSK

```{r, include=TRUE, echo=TRUE}

msk.protein.cnt <- msk %>%
  dplyr::inner_join(dbAlteration,by=c("Hugo_Symbol"="gene",
                                     "protein_change"="AAChange")) %>%
  dplyr::group_by(Hugo_Symbol,protein_change) %>%
  dplyr::summarize(n.patients.mutated=dplyr::n_distinct(PATIENT_ID),
                   source=paste0(unique(source),collapse=";")) %>%
  dplyr::arrange(desc(n.patients.mutated))
#print(dim(msk.protein.cnt))
#print(table(msk.protein.cnt$source,exclude = NULL))
knitr::kable(head(msk.protein.cnt))
```

### impose cancer type matches (protein)
```{r, include=TRUE}
msk.protein.cancer <- msk %>%
  dplyr::inner_join(dbRules,by=c("Hugo_Symbol"="gene",
                                  "protein_change"="AAChange",
                                 "CANCER_TYPE"="MskCancerType")) 
msk.protein.cancer.cnt <- msk.protein.cancer %>%
  dplyr::group_by(Hugo_Symbol,protein_change,CANCER_TYPE) %>%
  dplyr::summarize(n.patients.mutated=dplyr::n_distinct(PATIENT_ID),
                   source=paste0(unique(source),collapse=";"),
                   Drugs=paste0(unique(Drugs),collapse=";"),
                   ReferenceOrTrialID=paste0(unique(ReferenceOrTrialID),collapse=";")) %>%
  dplyr::arrange(desc(n.patients.mutated))
print(dim(msk.protein.cancer.cnt))
print(table(msk.protein.cancer.cnt$source,exclude = NULL))
print(sum(msk.protein.cancer.cnt$n.patients.mutated))
```

## MAF of cancer type specific actionabile variants 

```{r, include=TRUE, echo=TRUE}
### MAF results
maf.df.msk <- data.frame(MAF=msk$MAF)
maf.df.msk$source <- "MSK all variants"
maf.df.msk$actionability.summary <- "NA"

# maf of clinically actionable based on protein match
maf.df.protein<- data.frame(MAF=msk.protein.cancer$MAF)
maf.df.protein$source <- "clinically actionable"
maf.df.protein$actionability.summary <- msk.protein.cancer$actionability.summary

maf.df <- rbind(maf.df.msk,maf.df.protein)

outF <-  paste0(figDir,"/MAF_distribution_clinically_actionable.png")
ggplot(maf.df,aes(x=MAF,fill=source))+
  geom_histogram(alpha=.4)+
  theme_bw()+
  facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set1",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))
```

Now show density plot

```{r, include=TRUE, echo=TRUE}
ggplot(maf.df,aes(x=MAF,fill=source))+
  geom_density(alpha=.4)+
  theme_bw()+
  #facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set1",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))

```

```{r, include=TRUE, echo=TRUE}
ggplot(maf.df,aes(x=MAF,fill=actionability.summary))+
  geom_density(alpha=.4)+
  theme_bw()+
  #facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set1",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))

```

## MAF of actionabile variants in any cancer type
Do not constrain to cancer type

```{r, include=TRUE, echo=TRUE}
### MAF results
maf.df.msk <- data.frame(MAF=msk$MAF)
maf.df.msk$source <- "MSK all variants"
maf.df.msk$actionability.summary <- "NA"

# maf of clinically actionable based on protein match
maf.df.protein<- data.frame(MAF=msk.protein[!is.na(msk.protein$n.db.entries),"MAF"])
maf.df.protein$source <- "any clinically actionable"
maf.df.protein$actionability.summary <- msk.protein[!is.na(msk.protein$n.db.entries),"actionability.summary.mod"]

maf.df <- rbind(maf.df.msk,maf.df.protein)

ggplot(maf.df,aes(x=MAF,fill=source))+
  geom_histogram(alpha=.4)+
  theme_bw()+
  facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set2",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))
```

Now show density plot
```{r, include=TRUE, echo=TRUE}
ggplot(maf.df,aes(x=MAF,fill=source))+
  geom_density(alpha=.4)+
  theme_bw()+
  #facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set2",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))

```
```{r, include=TRUE, echo=TRUE}
ggplot(maf.df[!maf.df$actionability.summary =="Other",],aes(x=MAF,fill=actionability.summary))+
  geom_density(alpha=.4)+
  theme_bw()+
  #facet_grid(source~.,scale="free_y",)+
  xlab("MAF (%)")+
  ggtitle(paste0("MAF of clinically actioanable variants \n in MSK-IMPACT"))+
  scale_fill_brewer(palette="Set2",drop=FALSE)+
  theme(plot.title = element_text(hjust = 0.5))

```
